# Geospatial packages {.unnumbered}

Next week, we'll be working with maps and geospatial data. Working with this data can be somewhat difficult, and installing the packages can be a bit of a pain, so please make sure you have the following packages installed:

```{r}
#| eval=FALSE
install.packages("sf")
install.packages("ggspatial")
install.packages("rnaturalearth")
install.packages("rnaturalearthdata")
install.packages("wbstats")
install.packages("maps")


library(remotes)
remotes::install_github("ropensci/rnaturalearthhires")
```

If `sf` gives you trouble, you can try installing it from GitHub:

```{r}
#| eval=FALSE
install.package("remotes")
library(remotes)
install_github("r-spatial/sf")
```

Make sure everything loads correctly before we start.

```{r}
#| eval=FALSE
library(tidyverse)
library(sf)
library(ggspatial)
library(rnaturalearth)
library(wbstats)
library(maps)
```

If you're still having trouble, please consider switching to the class server for the next class. Especially on Macs, the `sf` package can be a bit problematic, and for some, requires the use of [homebrew](https://formulae.brew.sh/formula/sf), which should only be used if you know what you're doing.
